---
title: "Insurance Costs – Analysis of Factors Related to Costs"
author: "Isabel Holm"
date: "13 april 2026"
format: 
  html: 
    theme: lux
    fig-align: left
  pdf: default
editor: visual
# global options for rending
execute: 
  echo: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)

costs_raw <- read_csv("../data/insurance_costs.csv")

costs <- costs_raw |> 
  mutate(
    sex =               str_trim(sex),
    sex =               str_trim(sex),
    region =            str_trim(region),
    region =            str_to_title(region),
    smoker =            str_trim(smoker),
    smoker =            str_to_title(smoker),
    chronic_condition = str_trim(chronic_condition),
    chronic_condition = str_to_title(chronic_condition),
    exercise_level =    str_trim(exercise_level),
    exercise_level =    str_to_title(exercise_level),
    plan_type =         str_trim(plan_type),
    plan_type =         str_to_title(plan_type),
    sex =               as.factor(sex),
    region =            as.factor(region),
    smoker =            as.factor(smoker),
    chronic_condition = as.factor(chronic_condition),
    exercise_level =    as.factor(exercise_level),
    plan_type =         as.factor(plan_type)
  ) |> 
  mutate(
    age_group = case_when(
      age >= 18 & age <= 30 ~ "18-30",
      age >= 31 & age <= 40 ~ "31-40",
      age >= 41 & age <= 50 ~ "41-50",
      age >= 51 & age <= 60 ~ "51-60",
      age >= 61 & age <= 75 ~ "61-75",
      TRUE ~ NA_character_
    ),
    age_group = factor(
      age_group,
      levels = c("18-30", "31-40", "41-50", "51-60", "61-75"),
      ordered = TRUE
    )
  )
```

## Introduction

This report presents an analysis of insurance cost data with the aim of identifying which factors are most strongly associated with individual insurance charges. The dataset contains information on customer demographics, lifestyle factors, health-related variables, and insurance history.

The purpose of this analysis is twofold: first, to explore and describe patterns in the data through statistical summaries and visualizations; and second, to build
a regression model that can help explain variation in insurance costs and potentially serve as a support tool in future pricing decisions.

## Method

The analysis was conducted in R and structured across three main stages: data preparation, descriptive analysis, and regression modeling.

### Data Preparations

The dataset was loaded using the `readr` package and initially inspected using `glimpse()` and `summary()` to understand its structure and variable types.
Missing values were identified in three variables: **bmi** (28 missing), **exercise_level** (22 missing), and **annual_checkups** (20 missing). These were retained in the dataset and excluded on a case-by-case basis in calculations.

Categorical variables were standardized by trimming whitespace and applying consistent title case formatting, before being converted to factors. A new ordered variable, `age_group`, was created by binning age into five intervals (18–30, 31–40, 41–50, 51–60, 61–75) to support group-level comparisons in the descriptive analysis.

### Descriptive Analysis

Summary statistics for the target variable **charges** were computed, including mean, median, standard deviation, min, max, and interquartile range. A range of visualizations were also produced using `ggplot2` to explore the distribution of charges and its relationship with key variables. For the purpose of this report, only the most relevant findings are presented and discussed in the result section next.

## Results

## Statistical summary

```{r}
#| code-fold: true

costs |>
  summarise(
    n_charges       = n(),
    mean_charges    = mean(charges, na.rm = TRUE),
    median_charges  = median(charges, na.rm = TRUE),
    sd_charges      = sd(charges, na.rm = TRUE),
    min_charges     = min(charges, na.rm = TRUE),
    max_charges     = max(charges, na.rm = TRUE),
    iq_r_charges    = IQR(charges, na.rm = TRUE)
  ) |> 
  pivot_longer(
    cols      = everything(),
    names_to  = "Statistic",
    values_to = "Value"
  ) |> 
  knitr::kable(digits = 0)

```

The dataset contains 1,100 observations. Insurance charges range from 1,204 to 32,559, with a mean of 10,060 and a median of 9,124. The gap between the mean
and median suggests a right-skewed distribution, where a smaller number of high-charge customers pull the average upward. The standard deviation of 4,574
and an interquartile range of 5,873 indicate considerable spread in charges across the dataset.

Missing values present in **bmi,** **exercise_level,** and **annual_checkups** were excluded on a per-calculation basis using `na.rm = TRUE`, meaning all 1,100
observations are retained in the dataset and only omitted where the relevant variable is missing.

### Distribution of charges

```{r}
#| code-fold: true

ggplot(costs, aes(x = charges)) +
  geom_histogram(bins = 40, fill = "turquoise", color = "white") +
  labs(title = "Distribution of insurance charges", 
       x = "Charges",
       y = "Count")
```

The histogram confirms the right-skewed distribution noted in the summary statistics. The majority of customers have charges clustered between approximately 3,000 and 12,000, with a long tail extending toward the higher end. This suggests that while most customers incur moderate costs, a smaller group generates significantly higher charges, a pattern worth keeping in mind when interpreting the regression results.

### Charges per group

#### Charges by insurance plan type

```{r}
#| code-fold: true
#| label: Charges by insurance plan type

ggplot(costs, aes(x = plan_type, y = charges, fill = plan_type)) +
  geom_boxplot() +
  labs(title = "Charges by insurance plan type",
       x = NULL,
       y = "Charges") +
  theme(legend.position = "none")
```

Charges by insurance plan type show that Premium plan customers tend to have somewhat higher median charges compared to Basic and Standard customers. The Standard plan displays the widest spread and the most outliers at the upper end, suggesting greater variability in costs among that group. The difference between plan types is moderate, indicating that plan type alone may not be a strong standalone predictor of charges.

#### Charges by smoking status

```{r}
#| code-fold: true
#| label: Charges by smoking status

ggplot(costs, aes(x = smoker, y = charges, fill = smoker)) +
  geom_boxplot() +
  labs(title = "Charges by smoking status",
       x = NULL,
       y = "Charges") +
  theme(legend.position = "none")
```

Smoking status shows one of the clearest group differences in the dataset. Smokers have a considerably higher median charge of around 16,000 compared to approximately 8,000 for non-smokers. The interquartile range for smokers also sits entirely above the median for non-smokers, indicating that even typical smoker costs exceed what most non-smokers pay. This suggests smoking is likely to be one of the stronger predictors in the regression model.

#### Charges by chronic condition

```{r}
#| code-fold: true
#| label: Charges by chronic condition

ggplot(costs, aes(x = chronic_condition, y = charges, fill = chronic_condition)) +
  geom_boxplot() +
  labs(title = "Charges by chronic condition status",
       x = NULL,
       y = "Charges") +
  theme(legend.position = "none")

```

Customers with a chronic condition show notably higher charges than those without, with a median of around 13,000 compared to approximately 8,000 for the non-chronic group. The spread among chronic condition customers is also tighter, suggesting a more consistent cost pattern within that group. This points to chronic condition status as another likely significant predictor in the regression model.

#### Charges by age group

```{r}
#| code-fold: true
#| label: Charges by age group

ggplot(costs, aes(x = age_group, y = charges, fill = age_group)) +
  geom_boxplot() +
  labs(title = "Charges by age group",
       x = NULL,
       y = "Charges") +
  theme(legend.position = "none")
```

Charges show a clear upward trend across age groups, with median charges rising steadily from around 8,000 in the 18–30 group to approximately 12,000 in the 61-75 group. The spread also widens in older groups, particularly in the 61-75 bracket, suggesting greater variability in costs among older customers. This pattern aligns with the expectation that age is positively associated with insurance charges and indicates it will likely be a meaningful predictor in the regression model.

### Continuous predictors and charges

#### Age against charges, with smoking status

```{r}
#| code-fold: true
#| label: Age vs charges (with smoker status)

ggplot(costs, aes(x = age, y = charges, color = smoker)) +
  geom_point(alpha = 0.4) + 
  geom_smooth(method = "lm", se = FALSE, color = "tomato") +
  labs(title = "Age vs insurance charges (with smoking status)")
```

The scatter plot of age against charges, colored by smoking status, reveals two fairly distinct clusters. Smokers consistently occupy the upper portion of the chart across all age groups, while non-smokers cluster at lower charge levels. The overall trend line shows a positive relationship between age and charges, but the separation between smokers and non-smokers suggests that smoking status may have a stronger influence on charges than age alone.

#### Age against charges, with chronic condition

```{r}
#| code-fold: true
#| label: Age vs charges (with chronic condition status)

ggplot(costs, aes(x = age, y = charges, color = chronic_condition)) +
  geom_point(alpha = 0.4) + 
  geom_smooth(method = "lm", se = FALSE, color = "tomato") +
  labs(title = "Age vs insurance charges (with chronic condition)")

```

When grouping by chronic condition, the separation between groups is less pronounced than with smoking status. Customers with a chronic condition tend to have somewhat higher charges, but there is considerable overlap between the two groups. The positive relationship between age and charges holds regardless of chronic condition status.

## Regression analysis

........

## Conclusions 

.........

## Limitations

........

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

## Include a plot

```{r}
#| label: fig-Distribution of insurance charges-test
#| fig-cap: "A histogram of insurance charges"
#| fig-alt: "..more descriptive stuff here"
#| out-width: 50%
#| code-fold: true

ggplot(costs, aes(x = sex, y = charges, fill = sex)) +
  geom_boxplot() +
  labs(title = "Charges by gender",
       x = NULL,
       y = "Charges") +
  theme(legend.position = "none")

```
